Reasoning Graphically about Quantum Computation

### 2008-02-27 papers/reasoning-graphically/paper.tex
plugging of variable node graphs
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1376 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

The representation of Compact Closed Categories as Graphs, discuseed
in the previous section, is too restrictive for automated reasoning
about quantum computation. In particular, pictures such the Spider
Law, in Figure~\ref{spider-law-fig} are frequently needed. Such
pictures represent an infinite family of concrete graphs. In this
section we describe an extension of the representation for Compact
Closed Categories that allows a graph-based representation that
captures an infinite family of graphs. Notably, this allows us to
represent laws about quantum computation, such as the Spider
Law. Furthermore, this representation is still compositional in the
sense that we can plug graphs together. The extension, which we call
\ref{graph patterns}, adds two extra kinds of nodes to the
graph. Namely, {\em banged boxes} which informally represent an
arbitrary number of copies of a node (where the node itself can
contain a graph of the appropriate type), and {\em variable nodes}
which informally represent some, as yet unkown, node in a graph.

@@

In graph patterns, variable nodes generalise the concept of a boundary
node. Variable nodes, unlike boundary-nodes in the concrete graph, are
not classified into domain and co-domain. The intuition of a variable
node is that it can be replaced by any concrete node. Thus, unlike
nodes in conrete a graph, variable nodes do not define an ordering on
the edges that are incident. Instead, the edges that touch a variable
node are simply a set. The type of a variable nodes is thus defined as
a multiset rather than a tensor. Any tensor type can be lifted to a
multiset type in the obvious manner, which we denote by the operation
{\tt mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor
  b \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type, $T$,
is an instance of a multiset type, $M$, when it's lifting into a
multiset is a super-set of $M$: $ M \subseteq
\mathtt{mset\_of\_tensor}(T) \Longrightarrow T \in M$. The intuition
here is that an instance of a variable node is allowed to have as many
extra edges as it likes as long as it at least has the ones mentioned
in the original graph.

@@

A graph with variable nodes, $G$, can be given a formal semantics by
being interpretated as a set of concrete graphs which is denoted by
$\vinterp{G}$. We refer to members of this set as
instances. Formally, an instance of a graph with variable nodes is the
graph where every variable node has been replaced by some instance of
it. For example, in Figure~\ref{node-variable-instances-fig}, the
graphs $a$, $b$ and $c$ are instances of the graph $x$. Observe that
when a variable node has degree one, then it can be instantiated as a
boundary node in either the domain or co-domain.

A graph without boundary nodes but which may have variable nodes is
called a \emph{variable-node graph}. Such graphs are complete in the
sense that for every concrete graph there is a variable node graph
that can be instantiated to it: $\forall g.\,\exists G.\,g \in
\vinterp{G}$. The witness for this proof is made by simply replacing
all bounary nodes with variable nodes.

\subsection{Open Embedding}

\subsection{Plugging}

@@

@@

The key feature of rules like the Spider Law is that an arbirary
number of nodes can be matched. To support this we introduce an
operation, !-boxing, on graph representations. We draw !-boxes around
a set of chosen nodes and intuitively think of the resulting !-boxed
graph as representing a set of graphs with an arbirary number of
copies of the !-boxed nodes, where any connections to nodes outside
the !-box are also copied. The !-boxes can also be instantiated with
zero copies errasing all edges from the !-box. See
Figure~\ref{bang-box-ex1-fig} for a simple example.

A graph with !-boxes is simply a graph paired with a set of !-box node
sets, which define the !-boxes. The !-box node sets must either be
disjoint or strict subsets of each other. This is needed to avoid
confusion with overlaping boxes.

\subsubsection{Semantics.}

@@

\noindent The semantics of a !-boxed graph is then defined as the
members without any !-boxes of the set closed by these operation.  For
the purposes of this paper, the underlying graph representation of
!-boxed graphs is variable-node graphs. The set of variable-node
graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$.

\subsubsection{Matching}

\subsubsection{Plugging}

>>>>

The representation of Compact Closed Categories as Graphs, discuseed
in the previous section, is too restrictive for automated reasoning
about quantum computation. In particular, pictures such the Spider
Law, in Figure~\ref{spider-law-fig} are frequently needed. Such
pictures represent an infinite family of concrete graphs. In this
section we describe an extension to the representation that captures
an infinite family of graphs as a single finite graph. The extension,
which we call \ref{graph patterns}, adds two extra kinds of nodes to
the graph. Namely, {\em banged boxes} which informally represent an
arbitrary number of copies of a node (where the node itself can
contain a graph of the appropriate type), and {\em variable nodes}
which informally represent some, as yet unkown, node in a graph. We
also introduce a notion to replace graph composition, called graph
plugging that allows graph patterns to put together to form larger
graph patterns.

%Notably, this
%allows us to represent laws about quantum computation, such as the
%Spider Law. Furthermore, this representation is still compositional in
%the sense that we can plug graphs together.

@@

In graph patterns, variable nodes generalise the concept of a boundary
node. Variable nodes, unlike boundary-nodes in the concrete graph, are
not classified into domain and co-domain. The intuition of a variable
node is that it can be replaced by any subgraph of the right
type. Thus, unlike nodes in conrete a graph, variable nodes do not
define an ordering on the edges that are incident. Instead, the edges
that touch a variable node are simply a set. The type of a variable
nodes is thus defined as a multiset rather than a tensor. Any tensor
type can be lifted to a multiset type in the obvious manner, which we
denote by the operation {\tt mset\_of\_tensor}. For example, {\tt
  mset\_of\_tensor($a \tensor b \tensor a$) = $\{(a,2), (b,1)\}$}. We
say that a tensor-type, $T$, is an instance of a multiset type, $M$,
when it's lifting into a multiset is a super-set of $M$: $ M \subseteq
\mathtt{mset\_of\_tensor}(T) \Longrightarrow T \in M$. The intuition
here is that an instance of a variable node is allowed to have as many
extra edges as it likes as long as it at least has the ones mentioned
in the original graph.

@@

A graph with variable nodes, $G$, can be given a formal semantics by
being interpretated as a set of concrete graphs which is denoted by
$\vinterp{G}$. We refer to members of this set as instances. An
instance of a graph with variable nodes is the graph with every
variable node replaced by a subgraph of the right type. When a graph
has several varibale-nodes, the replacement subgraphs may be
intersecting. For example, in
Figure~\ref{node-variable-instances-fig}, the graphs $a$, $b$ and $c$
are instances of the graph $x$. A graph with variable nodes, and
without boundary nodes, is called a \emph{variable-node graph}. Such
graphs are complete in the sense that for every concrete graph there
is a variable node graph that can be instantiated to it: $\forall
g.\,\exists G.\,g \in \vinterp{G}$. The witness for this proof is made
by simply replacing all bounary nodes with variable nodes.

Below, we give a more formal and constuctive account the semnatics by
introducing a notion of an \emph{open embedding} and note that this
corresponds to the intuitive idea of graph-matching. We then go on and
introduce a way to join graphs with variable-nodes and note it's
relationship to composition on the underlying concrete graphs.

\subsection{Matching and Open Embedding}

\begin{definition}
  An \emph{open embedding} from $G$ to $H$ a strict embedding from the
  non-variable subgraph of $G$ into the non-variable subgraph of $H$
  and a mapping from the variable nodes in $G$ ...
\end{definition}

\subsection{Variable-node Plugging}

@@

\begin{definition}
  A \emph{plugging}, written $\pi_{p}(G,H)$, is a pair, $p$, of
  strict partial homomorpihc mappings, $f_G : G \Longrightarrow H$ and
  $f_H : H \Longrightarrow G$ such that $f_G$ and $f_H$ do not map
  concrete nodes and for the nodes that are mapped \emph{agree}
  following the definition given below.
\end{definition}

\begin{definition}
  Two mappings $f_G$ and $f_H$ \emph{agree} when $\forall v \in
  f_G(G).\, f_H(v) \in f^{-1}_G(v)$ and $\forall v \in f_H(H).\,
  f_G(v) \in f^{-1}_H(v)$. Intuitively this captures the idea a node
  is one function is always mapped to one the nodes that maps to it in
  the other function.
\end{definition}

We let $\pi(G,H)$ abbreviate $\exists p.\,\pi_p(G,H)$. We also
overload the notation $\pi_p(G,H)$ to let it denote the graph that
results from the plugging. This is defined to be the union of the
nodes and edges from $G$ and $H$ where the nodes and edges mapped by
the pair of plugging function are identified. Identification of nodes
instantiates variable nodes: when a variable node is mapped to a
concrete one, the result is a concrete node.

While this is simply the intuitive way to combine graphs with variable
nodes, the definition takes some care. In particular, it is important
that plugging preserves the types of concrete nodes and results in
subtypes in variable nodes. Figure~\ref{plugging-examples-fig}
presents some examples to help the reader get the intuition as well as
understand some of the more subtlte cases.

\begin{theorem}
  \emph{Plugging is symmetric}: $\forall p.\, \exists q.\, \pi_p(G,H)
  = \pi_q(H,G)$. Proof follows simply: given $p = (f_G,f_H)$, $q =
  (f_H,f_G)$. As a corrolary, we get $\pi(G,H) \Leftrightarrow
  \pi(H,G)$.
\end{theorem}

\begin{theorem}
  \emph{Match after plugging}: $G \leq \pi_f(G,H)$. And as a corrolary
  we get $H \leq \pi_f(G,H)$ by symmetry of plugging. 
\end{theorem}

\begin{theorem}
  \emph{Plugging preserves matching}: $V \leq G \Rightarrow V \leq
  \pi_f(G,H)$. As corrolaries we get $V \leq H \Rightarrow V \leq
  \pi_f(G,H)$ from symmetry of plugging and $\vinterp{G} \cap
  \vinterp{H} \subseteq \bigcup_f \vinterp{\pi_f(G,H)}$ from the
  definition of interpretation. 
\end{theorem}

@@

A crucial feature of rules like the Spider Law is that an arbirary
number of repetitions of a subgraph to be matched. To support this we
introduce an operation, !-boxing, on graph representations. Given a
graph representation, this introduces a new notation by outlining
(called !-boxing) a set of chosen nodes. Intuitively, the resulting
!-boxed graph can be thought of as representing a set of graphs with
an arbirary number of copies of the !-boxed nodes, where every copy
has the connects to the same nodes outside the !-box. The !-boxes can
also be instantiated with zero copies which is erases all edges from
the !-box. See Figure~\ref{bang-box-ex1-fig} for a simple example.

More formaly, a graph with !-boxes is simply a graph paired with a set
!-boxes. Each !-box is just the set of nodes inside it and each !-box
must be disjoint from the others.\footnote{The disjointness condition
  avoids a slightly more complex, although also more expressive,
  semantics for overlaping node sets in the !-boxes. While such
  expressivity is interesting from a representational point of view,
  it is not needed to express the currently known rules of quantum
  computation.}

For the purposes of this paper, the underlying graph representation of
!-boxed graphs is variable-node graphs. The set of variable-node
graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$. We
now define the semantics for !-boxed graphs, then show how to extend
an the embedding function which corresponds to matching as well as the
plugging operation for !-boxes.

\subsubsection{!-Box Semantics.}

@@

\noindent The semantics of a !-boxed graph is then defined as those
members of the set closed by these operation that do have any
!-boxes. Observe that an instance of a !-box graph is defined by
pairing each !-box with the natural number that defines how many
copies are made of it. Thus the $\binterp{G}$ is isomorphic to the set
of $k$-tuples of natural numbers, where $k$ is the number of !-boxes.

\subsubsection{!-Box Matching.}

Matching for !-box graphs is defined in terms of an underlying
matching operation. A !-box graph $G$ matches another one, $H$, iff
$\binterp{G} \supseteq \binterp{H}$. Matching can be given a more
constructive flavour by extending the open-embeddings of variable-node
graphs to !-boxes.

There is a !-open embedding of $G$ into $H$ if

\subsubsection{!-Box Plugging.}
